Fitness effects of mutations to SARS-CoV-2 proteins

### 2023-08-11 paper/paper_bioRxiv_v2.tex
manuscript formatted for final Virus Evolution submission (#42)
* made paragraph on last 5 kb of genome a bit more succinct

* add track changes manuscript

* change to author-date citations

* created separate PDF for supplemental material

* made version of `paper.tex` with refs in file and supp excluded for final submission

* make TIFF of fig1 and remove from final `paper.tex`

* make fig 2 a separate TIFF for final submission

* make fig 3 a separate EPS for final submission

* made fig 4 separate TIFF for final submission

* made fig 5 a separate EPS

* added paper with figs for early ac

The distributions of mutation effects concur with biological intuition about how different classes of mutations impact protein function.
Most synonymous mutations are nearly neutral, most stop codons are highly deleterious, and amino-acid mutations range from slightly beneficial to highly deleterious (Figure~\ref{fig:dist}A).
The handful of synonymous or noncoding mutations with highly deleterious effects are in either regions of known non-coding constraint (e.g., the ORF1ab ribosomal slippage site~\cite{bhatt2021structural}) or two regions in the center of E and the end of M (Figure~\ref{fig:synonymous}).

@@

The accessory proteins are encoded in the last 5kb of the SARS-CoV-2 genome, meaning that our results show that this portion of the genome (with the exception of the genes coding for nucleoprotein, E, and M) is under relatively weak purifying selection (Figure~\ref{fig:effects_dist_position_order}).
Note that some accessory proteins also vary in presence/absence across coronaviruses~\citep{v2021coronavirus,llanes2020betacoronavirus}, and are mostly dispensable in cell culture~\citep{mcgrath2022sars,silvas2021contribution,liu2022live,liu2014accessory}.
Decreased purifying selection in genes encoded near the end of the genome is also observed in more distantly related orders of large single-stranded RNA viruses such as closteroviruses~\citep{wang2011genetic, dawson2013virus}.

@@

We then performed quality control by ignoring any branch that met any of the following criteria:
\begin{itemize}
\item it had more than four nucleotide mutations;
\item it contained more than one nucleotide mutation that was a reversion to the Wuhan-Hu-1 reference sequence;
\item it contained more than one nucleotide mutation that was a reversion to the founder sequence for that clade as provided at \url{https://raw.githubusercontent.com/neherlab/SC2_variant_rates/7e738194a8c6592082f1caa9a6ca70cb68289790/data/clade_gts.json} by \cite{neher2022contributions};
\item it contained more than one nucleotide mutation to the same codon.
\end{itemize}
The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors or viral evolution in chronically infected humans, neither of which correspond to the pattern of typical SARS-CoV-2 transmission in acute infections.
Because the virus's evolution is very densely sampled, only a small fraction of branches have more than four mutations (Figure~\ref{fig:nmut_counts}).
The rationale for the second and third exclusions is that excess reversions can arise from base-calling pipelines that erroneously call low-coverage sites as reference.
We ignore branches with multiple nucleotide mutations to the same codon (this is very rare) because as detailed below our method is only designed to make estimates for mutations that represent single-nucleotide changes from the clade founder.
Note also that the mutation-annotated tree does not include insertion or deletion mutations, and so we only consider (and make estimates for) point mutations.

@@

\subsection{Comparison to dN/dS and other mutation-effect prediction algorithms}
For the comparison to the dN/dS approaches shown in Figure~\ref{fig:dnds}, we used the dN/dS values available at \url{https://github.com/spond/SARS-CoV-2-variation}~\cite{martin2021emergence} for all SARS-CoV-2 sequences, which were calculated using the FEL approach~\cite{kosakovsky2005not}.
Specifically, the file \url{https://raw.githubusercontent.com/spond/SARS-CoV-2-variation/master/windowed-sites-fel-all.csv} contains dN/dS estimates made using large-scale sequence sets from GISAID in three-month windows starting with the earliest sequences from December 2019 and continuing up until Jan-31-2022.
We averaged the dN/dS values for each site over all three month windows, and analyzed those time-window averaged dN/dS values.
In order to ensure comparability in the sequence sets used, for the comparisons in Figure~\ref{fig:dnds}, we used fitness estimates from our approach made using only sequences in the mutation-annotated tree as of Jan-31-2022.
We restricted our analysis to this timeframe because the large-scale dN/dS analyses at \url{https://github.com/spond/SARS-CoV-2-variation} (which use state-of-the-art methods) were only available for that range of dates.

The dN/dS ratios only provide a single number for each site, which cannot be directly compared to either the mutation-effect estimates or the deep mutational scanning, which estimate the effects of individual amino-acid mutations.
We therefore computed site-summary metrics of the mutation-effect estimates and the deep mutational scanning as the average effect of all measured amino-acid mutations at each site, excluding stop codons.
The correlations in Figure~\ref{fig:dnds} are with those site-summary metrics.

@@

\begin{figure*}[h]
\centering
\includegraphics[width=\linewidth]{figs/dnds.png}
\caption{
The site-mean of the estimated fitness effects outperforms dN/dS ratios as an indicator of functional constraint as assessed by correlation with deep mutational scanning measurements.
{\bf (A)}
Pearson correlation of log dN/dS ratio at each site with the mean estimated fitness of all amino acid mutations at each site from the current study.
For most genes, the log dN/dS ratio is correlated with the mean fitness effect, but the correlations are only modest (always substantially less than one).
The exception is the accessory proteins which are mostly under minimal amino-acid level selection (Figure~\ref{fig:dist}B) and only show a weak correlation.
{\bf (B)}
Correlation of the log dN/dS ratio or the mean estimated fitness effect at each site with the mean impact of amino-acid mutations at that site as experimentally measured by deep mutational scanning.
The numbers in the upper right of each panel give the Pearson correlation.
In every case, the mean estimated fitness effect (top row) is much better correlated with the experimental measurements than the log dN/dS (bottom row).
The dN/dS ratios are taken from \url{https://github.com/spond/SARS-CoV-2-variation}.
Note that the y-axis in each panel is extended so that it fully contains the trend line over the entire x-axis domain, explaining why the y-axis limits vary across the rows even though the range of the data are the same on the y-axis for each row.
Both the fitness estimates and dN/dS values shown here are calculated using only sequences available as of Jan-31-2022 (see Methods).
See \url{https://jbloomlab.github.io/SARS2-mut-fitness/public_2022-01-31/dnds_corr.html} for an interactive version of this plot, and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dnds_corr.html} for an interactive version that uses fitness estimates from all SARS-CoV-2 sequences including more recent ones.
\label{fig:dnds}
}
\end{figure*}

>>>>

The distributions of mutation effects concur with biological intuition about how different classes of mutations impact protein function.
Most synonymous mutations are nearly neutral, most stop codons are highly deleterious, and amino-acid mutations range from slightly beneficial to highly deleterious (Figure~\ref{fig:dist}A).
The handful of synonymous mutations with highly deleterious effects are in either regions of known non-coding constraint (e.g., the ORF1ab ribosomal slippage site~\cite{bhatt2021structural}) or two regions in the center of E and the end of M (Figure~\ref{fig:synonymous}).

@@

@@

We then performed quality control by ignoring any branch that met any of the following criteria:
\begin{itemize}
\item it had more than four nucleotide mutations;
\item it contained more than one nucleotide mutation that was a reversion to the Wuhan-Hu-1 reference sequence;
\item it contained more than one nucleotide mutation that was a reversion to the founder sequence for that clade as provided at \url{https://raw.githubusercontent.com/neherlab/SC2_variant_rates/7e738194a8c6592082f1caa9a6ca70cb68289790/data/clade_gts.json} by \cite{neher2022contributions};
\item it contained more than one nucleotide mutation to the same codon.
\end{itemize}
The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors of viral evolution in chronically infected humans, neither of which correspond to the pattern of typical SARS-CoV-2 transmission in acute infections.
Because the virus's evolution is very densely sampled, only a small fraction of branches have more than four mutations (Figure~\ref{fig:nmut_counts}).
The rationale for the second and third exclusions is that excess reversions can arise from base-calling pipelines that erroneously call low-coverage sites as reference.
We ignore branches with multiple nucleotide mutations to the same codon (this is very rare) because as detailed below our method is only designed to make estimates for mutations that represent single-nucleotide changes from the clade founder.
Note also that the mutation-annotated tree does not include insertion or deletion mutations, and so we only consider (and make estimates for) point mutations.

@@

\subsection{Comparison to dN/dS and other mutation-effect prediction algorithms}
For the comparison to the dN/dS approaches shown in Figure~\ref{fig:dnds}, we used the dN/dS values available at \url{https://github.com/spond/SARS-CoV-2-variation}~\cite{martin2021emergence} for all SARS-CoV-2 sequences, which were calculated using the FEL approach~\cite{kosakovsky2005not}.
The dN/dS ratios only provide a single number for each site, which cannot be directly compared to either the mutation-effect estimates or the deep mutational scanning, which estimate the effects of individual amino-acid mutations.
We therefore computed site-summary metrics of the mutation-effect estimates and the deep mutational scanning as the average effect of all measured amino-acid mutations at each site, excluding stop codons.
The correlations in Figure~\ref{fig:dnds} are with those site-summary metrics.

@@

\begin{figure*}[h]
\centering
\includegraphics[width=\linewidth]{figs/dnds.png}
\caption{
The site-mean of the estimated fitness effects outperforms dN/dS ratios as an indicator of functional constraint as assessed by correlation with deep mutational scanning measurements.
{\bf (A)}
Pearson correlation of log dN/dS ratio at each site with the mean estimated fitness of all amino acid mutations at each site from the current study.
For most genes, the log dN/dS ratio is correlated with the mean fitness effect, but the correlations are only modest (always substantially less than one).
The exception is the accessory proteins which are mostly under minimal amino-acid level selection (Figure~\ref{fig:dist}B) and only show a weak correlation.
{\bf (B)}
Correlation of the log dN/dS ratio or the mean estimated fitness effect at each site with the mean impact of amino-acid mutations at that site as experimentally measured by deep mutational scanning.
The numbers in the upper right of each panel give the Pearson correlation.
In every case, the mean estimated fitness effect (top row) is much better correlated with the experimental measurements than the log dN/dS (bottom row).
The dN/dS ratios are taken from \url{https://github.com/spond/SARS-CoV-2-variation}.
See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dnds_corr.html} for an interactive version of this plot.
\label{fig:dnds}
}
\end{figure*}
